If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-11-06 GroupPolarizationStatmod/paper/article-table.tex
updates

\rowcolor{gray!10}
\multirowcell{2}{Politics} 
                  & Moscovici \& Zavalloni (1969)~\cite{Moscovici1969} & The Group as a Polarizer of Attitudes & \emph{JPSP} \\
\rowcolor{gray!10}
                  & Schkade, Sunstein, \& Hastie (2010)~\cite{Schkade2010} & When Deliberation Produces Extremism & \emph{Critical~Review} \\[0.2em]

\rowcolor{gray!5}
\multirowcell{2}{Social Comp's \vspace{0.7em}}
% \multirowcell{2}{Social Comp's}
                  & Myers \& Bishop (1970)~\cite{Myers1970} & Discussion Effects on Racial Attitudes & \emph{Science} \\[0.2em]
\rowcolor{gray!5}
                  & Myers (1975)~\cite{Myers1975} & Discussion-Induced Attitude Polarization & \emph{Human Relations} \\[0.3em]

\rowcolor{gray!10}
\multirowcell{2}{Persuasive Arg's}
                  & Burnstein \& Vinokur (1973)~\cite{Burnstein1973} & Testing Two Classes of Theories About Group Induced Shifts In Individual Choice & \emph{JESP} \\
\rowcolor{gray!10}
                  & Burnstein \& Vinokur (1975)~\cite{Burnstein1975} & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments\ldots & \emph{JESP} \\

\rowcolor{gray!5}
\multirowcell{3}{Self-Categ'n}
                  & Abrams, et al., (1990)~\cite{Abrams1990} & Knowing What to Think by Knowing Who You Are: Self‐Categorization and the Nature of Norm Formation\ldots & \emph{Brit. J. of Soc. Psy.} \\
\rowcolor{gray!5}
                  & Hogg, Turner, \& Davidson (1990)~\cite{Hogg1990} & Polarized Norms and Social Frames of Reference: A Test of the Self-Categorization Theory of Group Polarization & \emph{B. \& Appl. Soc. Psy.} \\
\rowcolor{gray!5}
                  & Krizan \& Baron (2007)~\cite{Krizan2007} & Group Polarization and Choice Dilemmas: How Important is Self-Categorization? & \emph{Eur. J. of Soc. Psy.} \\

\rowcolor{gray!10}
\multirowcell{1}{Soc'l Decisions}
                  & Friedkin (1999)~\cite{Friedkin1999} & Choice Shifts \& Group Polarization & \emph{Amer. Soc. Rev.} \\

>>>>

\rowcolor{gray!5}
\multirowcell{2}{Politics} 
                  & Moscovici \& Zavalloni (1969)~\cite{Moscovici1969} & The Group as a Polarizer of Attitudes & \emph{JPSP} \\
\rowcolor{gray!5}
                  & Schkade, Sunstein, \& Hastie (2010)~\cite{Schkade2010} & When Deliberation Produces Extremism & \emph{Critical~Review} \\[0.2em]

\multirowcell{2}{Social Comp's \vspace{0.7em}}
% \multirowcell{2}{Social Comp's}
                  & Myers \& Bishop (1970)~\cite{Myers1970} & Discussion Effects on Racial Attitudes & \emph{Science} \\[0.2em]
                  & Myers (1975)~\cite{Myers1975} & Discussion-Induced Attitude Polarization & \emph{Human Relations} \\[0.3em]

\rowcolor{gray!5}
\multirowcell{2}{Persuasive Arg's}
                  & Burnstein \& Vinokur (1973)~\cite{Burnstein1973} & Testing Two Classes of Theories About Group Induced Shifts In Individual Choice & \emph{JESP} \\
\rowcolor{gray!5}
                  & Burnstein \& Vinokur (1975)~\cite{Burnstein1975} & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments\ldots & \emph{JESP} \\

\multirowcell{3}{Self-Categ'n}
                  & Abrams, et al., (1990)~\cite{Abrams1990} & Knowing What to Think by Knowing Who You Are: Self‐Categorization and the Nature of Norm Formation\ldots & \emph{Brit. J. of Soc. Psy.} \\
                  & Hogg, Turner, \& Davidson (1990)~\cite{Hogg1990} & Polarized Norms and Social Frames of Reference: A Test of the Self-Categorization Theory of Group Polarization & \emph{B. \& Appl. Soc. Psy.} \\
                  & Krizan \& Baron (2007)~\cite{Krizan2007} & Group Polarization and Choice Dilemmas: How Important is Self-Categorization? & \emph{Eur. J. of Soc. Psy.} \\

\rowcolor{gray!5}
\multirowcell{1}{Soc'l Decisions}
                  & Friedkin (1999)~\cite{Friedkin1999} & Choice Shifts \& Group Polarization & \emph{Amer. Soc. Rev.} \\

### 2025-11-06 GroupPolarizationStatmod/paper/intro.tex
updates

If an extremist moderates their opinion and it is measured with a Likert scale, will
this move to moderation be detected? The answer is often no in social psychology 
experiments~\cite{Liddell2018}. This
means that a group can appear to become radicalized over time even if the 
average opinion does not actually change~\cite{Turner2020}. 
This has huge consequences for the study of \emph{group polarization}, the hypothesis
that discussion among like-minded people pushes opinions toward extremes—effectively 
making it the science of ``echo chambers'', online or in-person communities 
composed of likeminded folks.  Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—but Roger Brown (1986, Ch. 5) wrote that group polarization ``does not occur with every group'' and that ``the
effect is not large,'' conflicting with the standard definition that lawlike
behavior means certain observations will obtain given that specific assumptions
are satisfied~\cite{Cartwright1999}. Given the prominence of group polarization 
theory, we decided to analyze whether its hundreds of claimed replications
may instead be measurement artifacts using a formal model and numerical simulations of
group polarization experiments.

@@

\subsection{Measurement procedure}

In most group polarization experiments, opinions are measured on an
ordinal scale: Burnstein and Vinokur (1975) for example used a ten-point
scale from 1 to 10 representing “the lowest odds of success acceptable” to
try a certain behavior, where 1 means 10\%—Moscovici and Zavalloni (1969) gave participants a seven-point scale, from “strongly disagree” (-3) to “strongly agree” (+3) to report their opinions. In these and every other survey-based group polarization study I know, the reported opinions on the ordinal scale—with discrete numbered bins—were treated as if they were continuous-valued, or in other words as if they were the internal mental opinion value itself, whatever that might be neurobiologically. 
Liddell and Kruschke (2018) showed that in the general case this setup can
lead to false positives because ordinal scales cannot track changes in
extreme opinions beyond the ceiling of the measurement scale—this approach is
ubiquitous in social psychology and survey research, not at all limited to
group polarization.

@@

We selected ten journal articles with sixty total experimental conditions
across the ten.

\begin{table}[ht]
\caption{\textbf{Journal articles in our corpus.}}
\centering
\begin{tabular}{p{1in}p{3.5in}p{2in}c}
\toprule
Authors (Year) & Title & Journal & Category \\
\midrule
  Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes &
    Journal of Personality and Social Psychology & Politics \\
    Schkade, Sunstein, \& Hastie & When Deliberation Produces Extremism & Critical Review
                                 & Politics \\
Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts
In Individual Choice & Journal of Experimental Social Psychology & Persuasive Arguments \\

Burnstein \& Vinokur (1975) & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments explanation of choice shifts & Journal of Experimental Social Psychology & Persuasive Arguments \\
\bottomrule
\end{tabular}
\label{table:mr}
\end{table}

>>>>

If an extremist moderates their opinion and it is measured with a Likert scale, will
this move to moderation be detected? The answer is often no in social psychology 
experiments~\cite{Liddell2018}. This
means that a group can appear to become radicalized over time even if the 
average opinion does not actually change~\cite{Turner2020}. 
This has huge consequences for the study of \emph{group polarization}, the hypothesis
that discussion among like-minded people pushes opinions toward extremes—effectively 
making it the science of ``echo chambers'', online or in-person communities 
composed of likeminded folks.  Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—a regular, quantitative relationship that
will always be observed when certain assumptions are
met~\cite{Cartwright1999}. Yet Roger Brown himself noted in Social Psychology
that group polarization ``does not occur with every group'' and that ``the
effect is not large''~\cite{Brown1986}. 
Given the influence of this theory and the hundreds of claimed ``replications'' of
group polarization, it is essential to ask whether any of these replications may
instead be \emph{measurement artifacts}—spurious effects created when the
``neutral'' point of the measurement scale fails to align with the group’s own
center of opinion. The science of opinion change has a central role in designing new
methods for social organization and social change to face pressing existential
challenges~\cite{Galesic2021,Galesic2023}.

@@

@@

We selected ten representative journal articles with sixty total experimental
conditions among them. For a representative sample of group polarization
research across decades, we two articles for three of the five article
categories: \emph{Social Comparisons}, \emph{Persuasive Arguments},
 and \emph{Politics}. We included three \emph{Self-Categorization} studies and one \emph{Social Decisions}.
All but the \emph{Politics} category refers to the theoretical tradition that the
authors were trying to prove right—the two \emph{Politics} articles were agnostic
to group psychology theories. This resulted in a representative sample research
over time, as well: we have one study from 1969,
four from the 1970s, three from the 1990s, one from 2007 and one from 2010. 
Information about our corpus is summarized below with references 
(Table~\ref{tab:articles}).

\include{article-table.tex}

Each study used subtly different experimental methods that make it difficult
to build cumulative general knowledge. But all of the studies share one thing
in common: they all measure participant opinions on an ordinal scale. We start
with a non-technical review of how ordinal opinion measurements work in group
polarization, and then how the distributions of measurements are analyzed to
determine if group opinions became polarized. Then we can explain our
close-reading analysis to extract the necessary experimental design parameters
we need to simulate all the experiments in our corpus.

The corpus contains
sixty different experimental conditions in the ten corpus journal articles. We 
simulate experiments for all 57 experimental conditions reported to have 
replicated group polarization, out of 60 total conditions (three were
null control conditions).

Using simulations we will estimate the number of these 57 replications that
must be retracted because null-polarization simulations appear polarized due
to measurement artifacts. We collect experiment design parameters that fill out 
details of the simulation model of group polarization experiments.

\subsection{Ordinal Opinion Measurements: Problems and Simulations}

In studies of group polarization, opinion measurements are typically ordinal,
reflecting ordered positions rather than measurable distances between them—
Moscovici and Zavalloni (1969) gave participants a seven-point scale, from
“strongly disagree” (-3) to “strongly agree” (+3) to report their opinions.
Burnstein and Vinokur (1975) used a ten-point scale from 1 to 10 representing
“the lowest odds of success acceptable” to try a certain behavior, where 1
means 10\%.  In these and every other survey-based group polarization study I
know, the reported opinions on the ordinal scale—with discrete numbered
bins—were treated as if they were continuous-valued, or in other words as if
they were the internal mental opinion value itself, whatever that might be
neurobiologically.

Note that ordinal scales clip all latent opinions with a magnitude greater
than the magnitude of its extreme values. This can make null-polarization
outcomes look like group polarization. Consider the following example 
with a group of just two people to see why: say they start with latent
opinions of 2.1 and 4.7, which they report as +2 and +3, let's say, on Moscovici
and Zavalloni's scale that tops out at +3. If they then discuss the topic and their
opinions move toward one another by 1 each, then their latent opinions would be 3.1
and 3.7. Before and after discussion the latent mean was a constant 3.4. 
When reported on an ordinal scale, both opinions are now +3—the average measured
opinion \emph{spuriously increased} from 2.75 to 3.0!

To determine whether a replication must be retracted, we developed a model to
simulate measurements of null-polarizations, labelled \emph{mere agreement}, where
the mean group opinion stays constant, but variance decreases. In other words,
simulated participants report more agreement following discussion, 
but they do not become radicalized.

The simulations are seeded by actual group polarization experiment parameters.
We copied the pre- and post-discussion opinion means reported in each article.
There is no original data available for any of the experiments. Sometimes more
information was available about the pre- and post-discussion
distributions—we used this information to decide whether replications plausibly 
have null-polarization. We recorded all available data with a guide to finding it in
its original article in the file `StudiesAnalysis.csv`, available with the
Supplemental Information or to view/download on
GitHub\footnote{\url{https://github.com/mt-digital/gp-statmod/GroupPolarizationStatmod/data/StudiesAnalysis.csv}}.

\subsection{Research Overview}

We collected original data across 10 journal articles to show that 95\% of 
the replications reported must be retracted due to non-identifiability with a
null-polarization model. These replications cannot be salvaged because the
original data is no longer available, and thus cannot be re-analyzed with a 
better statistical model that properly accounts for the measurement procedure.

In the Analysis we go further: we use simulations to show that 
the experimental designs in our corpus are beset by high false positive rates 
\emph{even with a better statistical model}. The only way to reduce false
positive rates with the experimental designs used in the papers is to increase
the significance threshold (in terms of Cohen's $d$ as explained in more
detail in the Methods).

Group polarization was taken as one of the most reliable group effects in social
science—but we showed it is impossible to distinguish 95\% of group
polarization detections from a null-polarization model. This means 54
replications must be retracted—and recall that this was a selection of the
highest quality, most representative group polarization research. 
To understand group polarization, new
methods will be required.

This sort of error threatens our ability to construct a
cumulative science of opinion change, social influence, and social change
generally. We need reliable social science as the foundation for new
strategies for social organization capable of reversing decline and powering
sustainable futures. We have a long way to go.

### 2025-11-06 GroupPolarizationStatmod/paper/method.tex
updates

\begin{table}[h]
  \caption{\textbf{Latent psychological variables and distributions.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $\omega_{i,t}$ & Latent opinion of individual $i$ at time $t$ & $\mathbb{R}$ \\
    $\mu_t$ & Mean latent opinion at time $t$ (constant in null-polarization) & $\mathbb{R}$ \\
    $\sigma_t$ & Variance of observed opnions at time $t$ & $\mathbb{R}$ \\
  \bottomrule
  \end{tabular} 
\end{table}

\begin{table}[h]
  \caption{\textbf{Experimental design and numerical model parameters.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $B$ & Number of bins & $5,6,7,\ldots$ \\
    $\theta_b$ & The $B+1$ threshold values that separate bins & 
      e.g., $-\infty, -2.5, -1.5,\ldots,2.5, \infty$ \\
    $b$ & Bin index—there are $B+1$ to mark $B$ bins & $0,1,2,\ldots,B$ \\
  \bottomrule
  \end{tabular} 
\end{table}

\begin{table}[h]
  \caption{\textbf{Simulated observed variables.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $\meanobst$ & Mean of all reported opinions (Eq. \ref{eq:meanobs}) & $\mathbb{R}$ \\
    $g$ & Observed group opinion polarization (Eq. \ref{eq:group-polarization})
        & $\mathbb{R}$ \\
    $d$ & Cohen's $d$ used as significance threshold 
        & $\mathbb{R}$ \\
    $\alpha_e,\alpha_s,\alpha_{\text{all}}$ & Family-wise/Type I error rate
    for an experiment, study, or across all experiments
             & $[0, 1]$ \\
    FDR & False discovery rate
        & $[0, 1]$ \\
  \bottomrule
  \end{tabular} 
\end{table}

>>>>

\input{method-tables.tex}
